---
title: "Adding SMDs to gtsummary tables with 3 or more groups"
author: "Zheer Kejlberg Al-Mashhadi"
date: "2023-11-04"
output: github_document
---

## Functions to add SMDs with explanations and examples of use
\

In this document, the necessary functions are defined and explained.
First, wrapper functions are defined for the core functionality (e.g., *core_smd_function* and *applied_smd_function*). Finally, four separate "caller" functions are created, each of which call the core functions in distinct ways for distinct use cases.
  
<br>
  
<br>
  
### Defining the functions
  
<br>
  
#### 1) Load dependencies

```{r, results='hide', warning=FALSE, message=FALSE, error=FALSE}

library(gtsummary) # For creating a baseline characteristics table
library(tidyverse) # For data wrangling and misc.

```
\

#### 2) Create the core functionality, via **core_smd_function()**, for taking the data and outputting the SMD results

```{r}

core_smd_function <- function(data, is_weighted, ref_group = FALSE) {
  # MAKE A TABLE OF EVERY POSSIBLE COMBO OF TWO DIFFERENT GROUPS
  groups <- factor(unique(data$by))
  pairs <- expand.grid(groups, groups) %>%
    arrange(as.integer(.data$Var1), as.integer(.data$Var2)) %>% 
    filter(Var1 != Var2) %>%
    filter(!duplicated(
      paste0(pmax(as.character(Var1), as.character(Var2)), 
             pmin(as.character(Var1), as.character(Var2)))))
  if (ref_group) { # IF ref_group, KEEP ONLY PAIRS CONTAINING THE REF GROUP
    pairs <- pairs %>% filter(Var1 == first(levels(groups)))
  }
  name_comparison <- function(pair) {
    filtered_data <- data %>% filter(by %in% pair) %>% mutate(by = factor(by))
    paste(levels(filtered_data$by)[1], levels(filtered_data$by)[2], sep = " vs. ")
  }
  comparisons <- apply(pairs, 1, name_comparison)
  
  # CREATE SUBSETS OF DATA
  subsetting <- function(pair, data) {
    as.data.frame(data) %>%
      filter(by %in% pair) %>%
      mutate(by = factor(by)) %>%
      droplevels()
  }
  
  data_subsets <- apply(X = pairs, MARGIN = 1, FUN = subsetting, data = data)
  
  # CALCULATE SMD BETWEEN GROUPS WITHIN EACH DATA SUBSET
  clean_SMD <- possibly(.f = smd::smd, otherwise = NA_real_) 
  
  if (is_weighted) { 
    smd_estimates <- map(data_subsets, ~ clean_SMD(.x$variable, .x$by, .x$weight_var))
  } else {
    smd_estimates <- map(data_subsets, ~ clean_SMD(.x$variable, .x$by)) 
  }
  
  extract_SMD <- possibly(.f = function(x){x$estimate}, otherwise = NA_real_)
  
  smd_estimates <- map_dbl(smd_estimates, ~ extract_SMD(.x))
  
  # OUTPUT THE RESULTS
  tibble(comp = comparisons, smd = smd_estimates) %>%
    spread(comp, smd) %>%
    relocate(any_of(comparisons))
}

```
\

#### 3) Create a function *clean_smd_data()* to prepare the input data for use by the *core_smd_function()*

```{r}

clean_smd_data <- function(data, variable, by, tbl) {
  data_type <- last(class(data))
  tbl_type <- first(class(tbl))
  if ((data_type != "survey.design" & data_type != "data.frame") | (tbl_type != "tbl_svysummary" & tbl_type != "tbl_summary")) {
    stop("Inappropriate input to smd function")
  }
  is_weighted <- data_type == "survey.design"
  
  if (is_weighted) {
    data <- data$variables %>% mutate(weight_var = 1 / data$allprob[[1]])
  } else {
    data <- data %>% mutate(weight_var = 1)
  }
  
  data <- dplyr::select(data, all_of(c(variable, by, "weight_var"))) %>%
    rlang::set_names(c("variable", "by", "weight_var")) %>%
    dplyr::filter(complete.cases(.))
  if (is.character(data$variable)) {
    data <- data %>% mutate(variable = factor(variable))
  }
  if (is.factor(data$variable)) {
    levels <- levels(data$variable)
  } else {
    levels <- NULL
  }
  return(list(data, levels, is_weighted))
}

```
\

#### 4) Create a function to apply the *core_smd_function()* in various ways for different use cases

This function first cleans the data and then applies the *core_smd_function()* but allows two distinct things to vary according to user preference:

1) The **location** argument (*"label"* or *"level"*):
- Specifying **"label"**, you get **one** SMD per variable. For categorical variables, a Mahalanobis distance is calculated between groups.
- Specifying **"level"**, you get an SMD *for each level* of all categorical variables. This option thus does not produce SMDs for continuous/numeric variables.
\
2) The **ref_group** argument (*TRUE* or *FALSE*):
- **FALSE**: There is no reference group, and SMDs will be calculated between every possible pair of groups (i.e., groups being defined by the "by" argument in *tbl_summary()* or *tbl_svysummary()*)
- **TRUE**: The first group (the first level of the variable given in the "by" argument - which is also the leftmost group in the table) will be set as a reference group, 

```{r}

applied_smd_function <- function(data, variable, by, tbl, ref_group = FALSE, location = "label") {
  clean_data <- clean_smd_data(data, variable, by, tbl)
  data <- clean_data[[1]]
  levels <- clean_data[[2]]
  is_weighted <- clean_data[[3]]
  
  if (location == "label") {
    output <- core_smd_function(data, is_weighted, ref_group = ref_group)
  } else { # location == "level"
    execute_by_level <- function(data, level, is_weighted) {
      data <- data %>% mutate(variable = variable == level)
      core_smd_function(data, is_weighted, ref_group = ref_group)
    }
    output <- map_dfr(levels, .f = ~ execute_by_level(data, .x, is_weighted))
  }
  return(output)
}

```
\

#### 5) Finally, four caller functions are defined to call the **applied_smd_function()** with preset values for *location* and for *ref_group*

```{r}

pairwise_smd <- function(data, variable, by, tbl, ...) {
  applied_smd_function(data, variable, by, tbl)
}
pairwise_smd_level <- function(data, variable, by, tbl, ...) {
  applied_smd_function(data, variable, by, tbl, location = "level")
}

focal_smd <- function(data, variable, by, tbl, ...) {
  applied_smd_function(data, variable, by, tbl, ref_group = TRUE)
}
focal_smd_level <- function(data, variable, by, tbl, ...) {
  applied_smd_function(data, variable, by, tbl, ref_group = TRUE, location = "level")
}

```
- **pairwise_smd** sets *ref_group = FALSE* and *location = "label"*, meaning all possible pairs of groups are compared, and one SMD per variable is returned.
- **pairwise_smd_level** sets *ref_group = TRUE* and *location = "label"*, meaning only group 1 is compared to all other groups, and one SMD per variable is returned.
- **focal_smd** sets *ref_group = FALSE* and *location = "level"*, meaning all possible pairs of groups are compared, and one SMD per level of cateogrical variables is returned.
- **focal_smd_level** sets *ref_group = TRUE* and *location = "level"*, meaning only group 1 is compared to all other groupsd, and one SMD per level of cateogrical variables is returned.
\
\

### Using the functions with gtsummary.
\

#### For unweighted data (with a *tbl_summary()* object), see the following examples:

For one SMD per variable, use either **pairwise_smd** or **focal_smd**. There's no need to specify an input to the location argument of add_stat (as this defaults to "label")
```{r, warning=FALSE, message=FALSE, error=FALSE}
trial %>% 
  tbl_summary(by = grade, include = c(trt, age, stage)) %>%
  add_stat(fns = everything() ~ pairwise_smd)
```

For one SMD per level of every categorical variable, use either **pairwise_smd_level** or **focal_smd_level** and specify *location = ~ "level"* in the *add_stat()* function
```{r, warning=FALSE, message=FALSE, error=FALSE}
trial %>% 
  tbl_summary(by = grade, include = c(trt, age, stage)) %>%
  add_stat(fns = everything() ~ pairwise_smd_level, location = ~ "level")
```
<center>
###### \*Notice, there are no SMDs for non-categorical variables due to the use of *pairwise_smd_level*
</center>
\

#### For weighted data, use *tbl_svysummary()*:

In this examply we use weights from **WeightIt** package. The **survey** package is necessary.
```{r, results='hide', warning=FALSE, message=FALSE, error=FALSE}

library(WeightIt) # To calculate weights
library(survey) # To create a surveydesign object (a "weighted" dataset)
```

Application of the *add_stat()* function is identical to the non-weighted case, but it is applied to a *tbl_svysummary* object instead of a *tbl_summary* object.
```{r, warning=FALSE, message=FALSE, error=FALSE}
trial %>% mutate(
  w = weightit(grade ~ age + stage + trt, data = .)$weights) %>% # create the weights
  survey::svydesign(~1, data = ., weights = ~w) %>% # creathe the svydesign object
  tbl_svysummary(by = grade, include = c(trt, age, stage)) %>%
  add_stat(fns = everything() ~ focal_smd)

```
<center>
###### \*Notice, comparisons are only made here between *group I* and all other groups due to the use of *focal_smd*
</center>